Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19

### 2024-07-20 manuscript/manuscript.qmd
Suparna edits

Did states misuse international emergency provisions during the COVID-19 crisis to justify human rights abuse or did they follow international law? Responding to the pandemic required extraordinary public health measures. Many governments restricted freedom of movement, association, and assembly, raising questions about states’ commitments to international human rights law. Some human rights treaties recognize states’ ability to use emergency measures during crises, and contain “escape clauses” that safeguard the permanent erosion of human rights during and following emergencies. The temporary suspension of state compliance from international treaties occurs through a legal action called derogation. However, the potential abuse of emergency measures during derogations raises concerns about their impact on human rights. Under emergency decrees, states often impose restrictions that are temporary in theory but difficult to roll back or contain in practice [@Stasavage2020; @LuhrmannRooney2021]. Did states use pandemic-era derogations how they were intended or were they used as a “pandemic pass” to justify violating human rights?

This research note explores state derogations and human rights behavior during the pandemic. Given recent medical advancements such as the distribution of vaccines globally and declining concern about COVID-19 severity, on May 5, 2023, the World Health Organization declared an end to the global health emergency. Many countries followed suit—the Biden administration, for instance, announced the end of United States’ emergency measures on May 11, 2023. We now assess important descriptive and analytical questions about states’ respect for international law and human rights during this emergency. This research note examines what derogating meant for compliance with international law during the pandemic.

Using data from the Varieties of Democracy PanDem dataset, the Oxford COVID-19 Government Response Tracker, and United Nations treaty derogation data, we ask: What is the impact of international legal emergency provisions on human rights?  How did derogating and non-derogating states vary in their implementation of emergency measures and human rights practices? We find that states that derogated and specified time limits to these emergency provisions generally stuck to the time limits and adhered to standards required during derogation by avoiding discriminatory policies and not violating non-derogable rights (the right to life, freedom from torture, and freedom from slavery). States that did not derogate from their international legal obligations and still imposed emergency measures were more likely to implement discriminatory policies, violate non-derogable rights, and have emergency measures without time limits.

This research note makes two contributions. First, it expands our understanding of compliance with flexibility mechanisms and derogations. While there is a significant literature on compliance with international law, especially human rights treaties [e.g. @Simmons:2009; @Cole:2012; @NielsenSimmons:2015; @ZvobgoSandholtzMulesky:2020; @Comstock:2021], and the flexibility tools such as reservations, understandings, and declarations (RUDs) that states invoke when they join or withdraw from such treaties [@Neumayer:2007; @Helfer:2005], we know little about states’ compliance with flexibility mechanisms such as derogations.  We explore how the potential decline in overall treaty compliance may actually be a good indication that states are using treaty design, and derogation option, as intended.

The pandemic further provides unique insight into derogation behavior because numerous states derogated under the same global conditions of a public health emergency, rather than domestic conflict or natural disaster, typical of treaty derogations, making it easier to compare state behavior during these derogations cross nationally. In making this comparison, this research note makes an important descriptive contribution showing how the implementation of policy measures differed across derogating and non-derogating states during the pandemic.

Second, this study contributes to the study of the relationship between international law and strategic repression by examining derogation behavior during a potentially tempting time for states to misuse them to increase repression and power during a time of global crisis. Previous derogation studies examine their use broadly, not during a defined global crisis. There is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Neumayer:2005; @Simmons:2009; @Merry:2009; @Hillebrecht:2014; @ConradRitter:2019]. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021]. Our findings are informative for international law, human rights, and repression. We find that derogating states during the pandemic did not increase abuse and violations. This supports the idea that states are being sincere about their intentions to comply with international law and not, generally, using actions like derogation to cover abusive behavior. This is an important finding for treaty design, including flexibility mechanisms, and the relationship between exploitation of international law to cover abuse.

Below, we first explore the purpose of derogations in international law, and then specifically during the pandemic. We focus on derogations to the International Covenant on Civil and Political Rights (ICCPR)—arguably the most significant and extensive treaty in the international human rights regime. We then introduce our expectations regarding how formally derogating from the treaty (or not) may impact a state’s human rights practices during the pandemic. We conclude with the implications of our research note for the study of international law during times of crises.

@@

Derogations in international law authorize states to temporarily suspend their treaty commitments, providing them with flexibility to respond to crises [@HBHelferFariss]. Times of crises can include wars, natural disasters, internal political unrest such as terrorism, and public health emergencies. When states derogate from their international legal obligations, it provides vital information to both international and domestic monitoring bodies, interest groups, and advocates about which rights are suspended, for how long, and the reasoning behind these suspensions. This information allows these actors—at least in principle—to challenge measures that are excessive, vague, or outlast the intended time frame of their implementation [@helfer2021].

Many legal assessments of derogations find importance in their use [@richardson2020], with some arguing that, “When a state does not bother with derogation provisions in times of emergencies where human rights are suspended, then it is a situation of unbridled state power” [@burchill2005, 97]. An abuse of state power can also potentially lead to the violation of non-derogable rights or rights whose protections cannot be affected by the declaration of a *de jure* emergency [@Greene:2020, 36]. Research supports this notion—there is no meaningful variation in states’ respect for derogable or non-derogable rights during officially declared states of emergencies [@richardsclay2012; @neumayer2013]. Even states that officially derogate from international treaties are more likely to violate physical integrity rights, which are non-derogable rights [@neumayer2013; @comstock2019].

@@

What kinds of states are more likely to submit derogations in the first place? Existing research focuses on the determinants of submitting derogations to the UN rather than the effects of these derogations. It finds that democracies are more likely to issue derogations and other post-commitment actions such as reservations, understandings, and declarations [@Simmons:2009; @comstock2019] and that most states facing an emergency do not issue derogations [@HBHelferFariss]. The former is because democracies typically need to convince domestic audiences of the legality and legitimacy of suspending rights during emergencies. Scholars were split on expectations about COVID-era derogations. @Lebret:2020 argued that derogations were the best legal mechanism for states to notify the global community of their inability to comply with human rights standards during the COVID-19 pandemic. However, @richardson2020 saw derogations as a less clear path for states during COVID-19, describing the actions and human rights as a “tangled morass” (p. 106).

@@

Between January 2020 and June 2021, `r derog_detail$countries` states issued `r derog_detail$total` ICCPR derogations (see @fig-derogation-map and @tbl-derogation-countries). The average length of a derogation was 41 days. States across multiple regions derogated, with notable concentration in South America, which has a robust history of engaging with human rights law. Notably, Nordic states—which typically are very active in human rights law engagement—did not issue any derogations during this period.

@@

The ICCPR is a foundational human rights treaty with broad human rights coverage. It is among a few human rights treaties that require states to report on the suspension or derogation of civil and political rights only under specific situations of “public emergency which threatens the life of the nation” [@iccpr]. The other treaties with this requirement include the European Convention on Human Rights (ECHR) and the American Convention on Human Rights (ACHR). Most of the states that derogated from the regional human rights conventions also derogated from the ICCPR, demonstrating that regional derogation activity did not preclude or replace ICCPR derogation activity.[^regional] The ICCPR is the only treaty with global membership---as of 2023, 173 states are party to the ICCPR. This means that most of the UN's 193 members had the ability to submit derogations. As such, we focus on ICCPR derogation activity in this research.[^future]

@@

To derogate, governments must first officially declare a state of emergency (Article 4). Derogations must fulfill three criteria: they should be temporary, proportional to the threat and non-discriminatory. Seven rights cannot be suspended or are non-derogable: the right to life, the prohibition of torture or cruel, inhumane punishment, the prohibition of slavery, the right to recognition as a person before the law, and the protection of freedom of thought, conscience and religion. However, several other ICCPR articles provide that public health needs can justify limitations on or derogations from certain rights [@richardson2020, 112]: freedom of movement (Article 12), freedom of expression (Article 19), and the rights to peaceful assembly and freedom of association (Article 22).

States follow different pathways to the actual submission of ICCPR derogations based on national level norms and policies. The UN does not specify the set pathway for national derogation approval. Most states followed a pathway of executive level instruction and authorization. For example, Argentina's derogations were described, upon submission to the UN as following an emergency decree issued by its executive branch [@un-derog-arg]. Guatemala directly referenced the will of the President of the state in initiating the derogation policy [@un-derog-gtm]. Some states such as the Dominican Republic, specified the role of the legislature in advancing the derogation provision [@un-derog-dom]. Other states, such as Azerbaijan, referenced how the executive and legislature were involved in approving [@un-derog-aze]. In reading through the written derogations, eighteen of the issuing states indicated that the executive or head of state directed the derogation, and five indicated the national legislature's role in approving it. All written derogations referenced a national declaration of emergency.

Almost all states that issued derogations specified time limits, as the derogation format requires. Fifteen of the derogating states issued derogations that extended or renewed prior ones. The frequency of renewal signifies that states took the times specified within the derogations seriously and updated the treaty action when their national government expected the times of crises to continue. We understand the time specification to be a high level of precision about obligation, applying the dimensions of legalization [@AbbottKeohaneMoravcsik:2000] which contributes to a higher expectation of compliance, or adherence with the treaty.

@@

> **Hypothesis 1~a~**: If states issued COVID-19 derogations, then they will be more likely to impose emergency policies that cancel public events, restrict gatherings, close public transit, and restrict movement.
>
> **Hypothesis 1~b~**: If states did not issue COVID-19 derogations, then they will be less likely to impose relevant emergency policies.

@@

If states determine that emergency measures are justified without regard for international law, the processes for respecting rights during crises and restoring rights at the end of the emergency may not be followed. Many states used the pandemic as an excuse to centralize authority. Governments engaged in discriminatory and abusive policy enforcement—they imposed restrictions on travel or movement by minority or opposition groups. For instance, in Australia, indigenous and migrant communities were disproportionately targeted in the enforcement of movement restrictions [@ICRC2021]. In India, the police used a ban on public gatherings as an excuse to get rid of a months-long sit-in protesting the country’s new citizenship law seen as discriminatory towards Muslims [@bbc2020].

The pandemic also led to the violation of non-derogable rights in many states. For instance, Greek officials intercepted and turned back boats filled with asylum-seekers and summarily returned asylum-seekers at the land border with Turkey. In doing so, they violated the non-derogable principle of non-refoulement—Article 7 of ICCPR prohibits people being sent back to another state where they are in danger of ill-treatment or torture [@HathawayStevensLim:2020]. Many other states changed rules of confinement affecting detainees, which also potentially violated non-derogable rights. For instance, in March 2020, a French government order extended the duration of pre-trial detention, raising the issue of prison overcrowding and leading to inhumane and degrading treatment [@Lebret:2020]. Besides violating a non-derogable right, this also raised questions regarding the proportionality of emergency measures.

We expect that states that did not formally derogate and did not notify the ICCPR of impending rights suspensions were less likely to take the criteria for derogations—necessity, proportionality, non-discrimination and time delimited—seriously, and were therefore more likely to violate these criteria.

> **Hypothesis 2~a~**: If states formally derogate from international treaty provisions, then they are less likely to use discriminatory policy measures during the pandemic.
>
> **Hypothesis 2~b~**: If states formally derogate from international treaty provisions, then they are less likely to violate non-derogable rights.
>
> **Hypothesis 2~c~**: If states formally derogate from international treaty provisions, then they are less likely to impose measures with no time limits.

@@

Next, we explore the relationship between formal treaty derogations on four different human rights and policy outcomes from V-Dem's PanDem project [@vdem-pandem]: (1) whether pandemic-related emergency measures violated the ICCPR's principle of non-discrimination, (2) whether countries derogated from non-derogable ICCPR rights, (3) whether emergency measures failed to state a time limit, and (4) whether state security forces violated physical integrity rights while enforcing emergency measures. Similar to the Oxford data, each of these outcomes is an ordered category measuring the severity of the outcomes, including no, minor, moderate, and major violations. We do not collapse these variables into binary indicators and instead work with them as ordered categories.

@@

We use a binary indicator of whether a state was under ICCPR derogation conditions during a given week. We collected and coded this data from real time state action updates recorded at the UN Treaty Collection [@UnitedNations:2023]. During the 15 months covered in this project, there were 115 formal derogations submitted mentioning 8 specific ICCPR articles; 15 derogations made no reference to any specific article. Of the derogations that indicated specific time limits, the average derogation duration was 41 days. Fifteen of the derogating states issued renewals/extensions of derogations that had time limits. We mark the start and end date of each derogation so that each country-week observation indicates whether a state has derogated.

@@

Several additional variables that are associated with and potentially confound both the decision to derogate and the decision to enact specific policies or human rights-related actions. To account for the severity of the pandemic in each country, we control for the count of new COVID-19 cases and the count of new COVID-19 deaths, as well as the cumulative count of cases and deaths.

@@

When working with binary outcomes like the declaration of general emergency measures or the violation of non-derogable rights, we use logistic regression models; when working with ordered categorical outcomes like the degree of discriminatory policy or abusive enforcement, we use ordered logistic regression models. Because the the panel data represents a complete or apparent sample of all countries [@BerkWesternWeiss:1995], we use a Bayesian approach to model the probabilities of the various outcomes and better estimate the uncertainty in the model parameters. We use weakly informative priors [@GelmanJakulinPittau:2008] for all model coefficients to keep estimates within realistic ranges and allow for reasonable uncertainty in parameter estimates.

@@

[^fixed-note]: In these formulas, we refer to all the $\beta$ coefficients as "population averages and coefficients." The nomenclature for these terms is inconsistent across disciplines, though, and they are also often referred to as "global effects" or "fixed effects". Broadly speaking, these $\beta$ terms represent the effects of each variable on the outcome across all regions or countries and do not incorporate region- or country-specific differences.

@@

We first examine how derogating and non-derogating states implemented emergency public health measures. @fig-preds-mfx-policies-1 shows the results of five logistic regression models as predicted probabilities of imposing specific emergency measures over the course of the first 15 months of the pandemic. Countries with a derogation in effect are more likely to have undertook each policy, and the difference in probabilities between derogating and non-derogating states is substantial (see @fig-preds-mfx-policies-2 for the percentage-point differences between the predictions for the two groups of states; see Appendix Table 1 for numeric values). This provides strong support for both components of our first hypothesis—states that derogated were substantially more likely to impose relevant emergency policies aimed at curtailing the spread of the virus (H~1a~), while states that did not issue derogations were less likely to engage in similar policies (H~1b~).

All countries were highly likely to cancel public events in March and April 2020, with a greater than a 95% chance. Over time, countries that did not derogate became less likely to cancel events, with the probability dropping to just above 90% in June 2021. Countries that derogated, however, regularly canceled with a nearly 100% probability. The difference in predicted probabilities between the two types of countries ranged from `r policies_mfx$cancel_public_events.min$draw_nice`–`r policies_mfx$cancel_public_events.max$draw_nice` percentage points (`r policies_mfx$cancel_public_events.max$p_gt`) between March 2020 and June 2021. The differences in policy implementation are starker for other types of emergency measures. Derogating states had a nearly 100% probability of imposing restrictions on gathering across all 15 months, while non-derogating states ranged between `r policies_pred$gathering_restrictions.No.min$draw_nice`% and `r policies_pred$gathering_restrictions.No.max$draw_nice`%, representing a `r policies_mfx$gathering_restrictions.min$draw_nice`–`r policies_mfx$gathering_restrictions.max$draw_nice` percentage point difference (`r policies_mfx$gathering_restrictions.max$p_gt`). Derogating states were also far more likely to close public transit, beginning with a `r policies_pred$close_public_transit.Yes.max$draw_nice`% probability in March 2020 and ending in June 2021 with a `r policies_pred$close_public_transit.Yes.min$draw_nice`% probability. Non-derogating states followed a similar downward trend over time, but `r policies_mfx$close_public_transit.min$draw_nice`–`r policies_mfx$close_public_transit.max$draw_nice` percentage points lower than their derogating counterparts (`r policies_mfx$close_public_transit.max$p_gt`). Derogating states were more likely to restrict movement, with a nearly 100% probability, dropping to `r policies_pred$movement.Yes.min$draw_nice`% by June 2021. Non-derogating states began the pandemic with a predicted `r policies_pred$movement.No.max$draw_nice`% probability of limiting movement, dropping to only `r policies_pred$movement.No.min$draw_nice`% by June 2021, representing a `r policies_mfx$movement.min$draw_nice`–`r policies_mfx$movement.max$draw_nice` percentage point difference (`r policies_mfx$movement.max$p_gt`). The only policy where there was no sizable difference between the two types of states is restrictions on international travel. Both derogating and non-derogating states had a nearly 100% probability of limiting international movement across all 15 months.

@@

These results show that countries implementing emergency measures seem to have taken the time to follow international law and uphold their treaty obligations. International law appears to not just be an empty commitment—states that derogated were more likely to keep these emergency measures in place for longer and take their derogations seriously.

>>>>

Did states misuse international emergency provisions during the COVID-19 crisis to justify human rights abuse or did they follow international law? Responding to the pandemic required extraordinary public health measures. Many governments restricted freedom of movement, association, and assembly, raising questions about states' commitments to international human rights law. Some human rights treaties recognize states' need for emergency measures during crises, and contain “escape clauses” that safeguard the permanent erosion of rights during emergencies. The temporary suspension of state compliance from international treaties occurs through a legal action called derogation. However, the potential abuse of measures during derogations raises concerns about their impact on human rights. Under emergency decrees, states often impose restrictions that are temporary in theory but difficult to roll back or contain in practice [@Stasavage2020; @LuhrmannRooney2021]. Did states use pandemic-era derogations how they were intended or were they used as a “pandemic pass” to justify violating human rights?

Given recent medical advancements such as the distribution of vaccines globally and declining concern about COVID-19 severity, we assess important descriptive and analytical questions about states' respect for international law and human rights during this emergency. This research note examines what derogating meant for compliance with international law during the pandemic. Looking at the use of derogations during a public health emergency is especially useful as the countries that derogated from the main international human rights treaty—the International Covenant on Civil and Political Rights (ICCPR)—do not appear to coincide with much of the conventional wisdom about which states are mostly likely to abide by international human rights treaty commitments. While countries with strong democracies and presence of international NGOs [@Neumayer:2005], or high state capacity and bureaucratic efficacy [@Cole:2015], or strong domestic and judicial institutions [@Hathaway:2002; @Simmons:2009; @PowellStaton:2009] are typically more likely to comply with human rights treaties, @tbl-derogation-countries shows that this was not exactly the set of countries that utilized derogations during the pandemic. Moreover, even these derogation patterns do not align with previous research that finds that democracies are most likely to derogate [@HBHelferFariss]. These descriptive statistics highlight the uniqueness in the conditions under which states comply with their international human rights treaty commitments and when they decide to derogate from them, especially in public health emergencies. This question is thus important for the study of international law, compliance, and human rights during times of crises.

Using data from the Varieties of Democracy PanDem dataset, the Oxford COVID-19 Government Response Tracker, and United Nations treaty derogation data, we ask: What is the impact of international legal emergency provisions on human rights? How did derogating and non-derogating states vary in their implementation of emergency measures and human rights practices? We find that states that derogated generally stuck to the time limits of emergency measures and adhered to required standards by avoiding discriminatory policies and not violating non-derogable rights (the right to life, freedom from torture, and freedom from slavery). States that did not derogate and still imposed emergency measures were more likely to implement discriminatory policies, violate non-derogable rights, and have emergency measures without time limits.

This research note makes three contributions. First, it expands our understanding of compliance with flexibility mechanisms and derogations. While there is a significant literature on compliance with international law, especially human rights treaties [e.g. @Simmons:2009; @Cole:2012; @NielsenSimmons:2015; @ZvobgoSandholtzMulesky:2020; @Comstock:2021], and the flexibility tools such as reservations, understandings, and declarations (RUDs) that states invoke when they join or withdraw from such treaties [@Neumayer:2007; @Helfer:2005], we know much less about states' compliance with flexibility mechanisms such as derogations. However, compliance with derogations may not necessarily be the same as compliance with treaty obligations. Treaty compliance is often motivated by international-level factors or concerns such as reciprocity, retaliation, signaling, and reputations [@SimmonsHopkins:2005; @Hathaway:2007; @Hafner-BurtonLeVeckVictor:2017]. However, countries derogate primarily for domestic political or institutional reasons [@HBHelferFariss], and international-level factors were relatively unimportant during the pandemic where preventing the domestic transmission of a virus and restricting citizens' rights to this end were governments' main motivations to derogate or not. As such, compliance with derogations is distinct from compliance with treaties and the former is relevant to our understanding of the latter—without derogations, states may repress more rights during crises, including non-derogable rights. Analyzing the behavior of derogating states and if it systematically differs from those that do not has important implications for the design of international treaties and whether derogations actually provide the safety valves that governments need during crises.

Second, this study contributes to the study of the relationship between international law and strategic repression by examining derogation behavior during a potentially tempting time for states to misuse them to increase repression and power during a time of global crisis. There is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Neumayer:2005; @SimmonsHopkins:2005; @Merry:2009; @Hillebrecht:2014; @ConradRitter:2019]. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021]. We find that overall these concerns are overstated, and derogating states during the pandemic did not increase abuse and violations. This supports the idea that states are being sincere about their intentions to comply with international law and not, generally, using actions like derogation to cover abusive behavior. This is an important finding for the relationship between treaty design exploitation of international law to cover abuse.

Finally, this research note looks at an understudied area in which derogations happen—public health emergencies. Derogations during the pandemic do not appear to hold the same patterns as found in previous literature according to which stable democracies and countries where domestic courts can exercise strong oversight of the executive are most likely to derogate. In addition, democracies are the most serial derogators in response to grave external threats such as terrorism [@HBHelferFariss]. However, as @tbl-derogation-countries shows, this was not the case during the pandemic. The pandemic further provides unique insight into derogation behavior because numerous states derogated under the same global conditions of a public health emergency, rather than domestic conflict or natural disaster, typical of treaty derogations, making it easier to compare state behavior during these derogations cross nationally. Thus, this research note makes an important descriptive contribution showing how the implementation of policy measures differed across derogating and non-derogating states during the pandemic.

Below, we first explore the purpose of derogations in international law, and then specifically during the pandemic. We focus on derogations to the International Covenant on Civil and Political Rights (ICCPR)—arguably the most significant and extensive treaty in the international human rights regime. We then introduce our expectations regarding how formally derogating from the treaty (or not) may impact a state's human rights practices during the pandemic. We conclude with the implications of our research note for the study of international law during times of crises.

@@

Derogations in international law authorize states to temporarily suspend their treaty commitments, providing them with flexibility to respond to crises [@HBHelferFariss]. These crises can include wars, natural disasters, internal political unrest such as terrorism, and public health emergencies. When states derogate from their international legal obligations, it provides vital information to both international and domestic monitoring bodies, interest groups, and advocates about which rights are suspended, for how long, and the reasoning behind these suspensions. This information allows these actors—at least in principle—to challenge measures that are excessive, vague, or outlast the intended time frame of their implementation [@helfer2021].

Many legal assessments of derogations find importance in their use [@richardson2020], with some arguing that, “When a state does not bother with derogation provisions in times of emergencies where human rights are suspended, then it is a situation of unbridled state power” [@burchill2005, 97]. An abuse of state power can also potentially lead to the violation of non-derogable rights [@Greene:2020, 36]. Research supports this notion—there is no meaningful variation in states' respect for derogable or non-derogable rights during officially declared states of emergencies [@richardsclay2012; @neumayer2013]. Even states that derogate from international treaties are more likely to violate physical integrity rights, which are non-derogable rights [@neumayer2013; @comstock2019].

@@

What kinds of states are more likely to submit derogations in the first place? Existing research focuses on the determinants of submitting derogations to the UN rather than the effects of these derogations. Democracies are more likely to issue derogations and other post-commitment actions such as reservations, understandings, and declarations [@Simmons:2009; @comstock2019]; most states facing an emergency do not issue derogations [@HBHelferFariss]. The former is because democracies typically need to convince domestic audiences of the legality and legitimacy of suspending rights during emergencies. Scholars were split on expectations about COVID-era derogations. @Lebret:2020 argued that derogations were the best legal mechanism for states to notify the global community of their inability to comply with human rights standards during the COVID-19 pandemic. However, @richardson2020 saw derogations as a less clear path for states during COVID-19, describing the actions and human rights as a “tangled morass” (p. 106).

@@

Between January 2020 and June 2021, we find that `r derog_detail$countries` states issued `r derog_detail$total` ICCPR derogations (see @fig-derogation-map and @tbl-derogation-countries). The average length of a derogation was 41 days. States across multiple regions derogated, with notable concentration in South America, which has a robust history of engaging with human rights law. Notably, Nordic states—which typically are very active in human rights law engagement—did not issue any derogations.

@@

The ICCPR is a foundational human rights treaty and among the few treaties that requires states to report on the suspension or derogation of civil and political rights only under specific situations of “public emergency which threatens the life of the nation” [@iccpr]. The other treaties with this requirement include the European Convention on Human Rights (ECHR) and the American Convention on Human Rights (ACHR). Most of the states that derogated from the regional human rights conventions also derogated from the ICCPR, demonstrating that regional derogation activity did not preclude or replace ICCPR derogation activity.[^regional] The ICCPR is the only treaty with global membership---as of 2023, 173 states are party to the ICCPR. Thus, most of the UN's 193 members had the ability to submit derogations and we focus on ICCPR derogation activity in this research.[^future]

@@

To derogate, governments must first declare a state of emergency (Article 4). Derogations must fulfill three criteria: they should be temporary, proportional to the threat and non-discriminatory. Seven rights cannot be suspended or are non-derogable: the right to life, the prohibition of torture or cruel, inhumane punishment, the prohibition of slavery, the right to recognition as a person before the law, and the protection of freedom of thought, conscience and religion. However, several other ICCPR articles provide that public health needs can justify limitations on or derogations from certain rights [@richardson2020, 112]: freedom of movement (Article 12), freedom of expression (Article 19), and the rights to peaceful assembly and freedom of association (Article 22).

States follow different pathways to the actual submission of ICCPR derogations based on national level norms and policies; the UN does not specify a uniform procedure. Most states followed a pathway of executive level instruction and authorization. For example, Argentina's derogations were described, upon submission to the UN, as following an emergency decree issued by its executive branch [@un-derog-arg]. Guatemala directly referenced the will of the President of the state in initiating the derogation policy [@un-derog-gtm]. Some states such as the Dominican Republic, specified the role of the legislature in advancing the derogation provision [@un-derog-dom]. Others, such as Azerbaijan, referenced how the executive and legislature were involved in approving [@un-derog-aze]. In reading through the written derogations, eighteen of the issuing states indicated that the executive or head of state directed the derogation, and five indicated the national legislature's role in approving it. All written derogations referenced a national declaration of emergency.

Almost all states that issued derogations specified time limits. Fifteen states extended or renewed prior derogations. The frequency of renewal signifies that states took the time specified within the derogation seriously and updated the treaty action when their national government expected the times of crises to continue. We understand the time specification to be a high level of precision about obligation, applying the dimensions of legalization [@AbbottKeohaneMoravcsik:2000] which contributes to a higher expectation of compliance, or adherence with the treaty.

@@

> **Hypothesis 1~a~**: If states derogate from the ICCPR, then they will be more likely to impose emergency policies that cancel public events, restrict gatherings, close public transit, and restrict movement.
>
> **Hypothesis 1~b~**: If states did not derogate from the ICCPR, then they will be less likely to impose relevant emergency policies.

@@

If states determine that emergency measures are justified without regard for international law, the processes for respecting rights during crises and restoring rights at the end of the emergency may not be followed. Many states used the pandemic as an excuse to centralize authority. Governments engaged in discriminatory and abusive policy enforcement—they imposed restrictions on travel or movement by minority or opposition groups. For instance, in Australia, indigenous and migrant communities were disproportionately targeted in the enforcement of movement restrictions [@ICRC2021]. In India, the police used a ban on public gatherings as an excuse to get rid of a months-long sit-in protesting the country's new citizenship law seen as discriminatory towards Muslims [@bbc2020].

The pandemic also led to the violation of non-derogable rights in many states. For instance, Greek officials intercepted and turned back boats filled with asylum-seekers and summarily returned asylum-seekers at the land border with Turkey. In doing so, they violated the non-derogable principle of non-refoulement—Article 7 of ICCPR prohibits people being sent back to another state where they are in danger of ill-treatment or torture [@HathawayStevensLim:2020]. Many other states changed rules of confinement affecting detainees, which also potentially violated non-derogable rights. For instance, in March 2020, a French government order extended the duration of pre-trial detention, raising the issue of prison overcrowding and leading to inhumane and degrading treatment [@Lebret:2020].

We expect that states that did not formally derogate and did not notify the ICCPR of impending rights suspensions were less likely to take the criteria for derogations—necessity, proportionality, non-discrimination and time delimited—seriously, and were therefore more likely to violate these criteria. However, we posit that derogations may be associated with fewer discriminatory policies due to two potential mechanisms. First, derogating states may be more sincere about their treaty commitments, which may have motivated them to derogate in the first place. Second, states that derogate also invite increased domestic and international monitoring, leading to more rights-respecting behavior. The UN Human Rights Committee (UNHRC) reviews derogations, may challenge them, and also examines complaints filed by individuals.[^individuals] It can also lead to involvement of domestic stakeholders in the process; their monitoring can lead to increased incentive for the government to make sure they are not violating the terms of the derogations. For instance, after Armenia derogated from the ICCPR and the ECHR, civil society and media representatives provided feedback on emergency measures affecting them—a move that was welcomed both by the OSCE Media Freedom Representative and the Armenian media [@OSCE:2020]. Further, for regional treaties not examined in this paper, such as the ECHR, individuals can also seek recourse at the European Court of Human Rights. Thus, there could be sincere as well as strategic reasons derogating states may behave differently from non-derogating states.

[^individuals]: Challenges to derogations by individuals and as part of the review of state reports is quite common [@Nowak:2005].

> **Hypothesis 2~a~**: If states derogate from the ICCPR, then they are less likely to use discriminatory policy measures during the pandemic.
>
> **Hypothesis 2~b~**: If states derogate from the ICCPR, then they are less likely to violate non-derogable rights.
>
> **Hypothesis 2~c~**: If states derogate from the ICCPR, then they are less likely to impose measures with no time limits.

@@

Next, we explore the relationship between formal treaty derogations and four different human rights and policy outcomes from V-Dem's PanDem project [@vdem-pandem]: (1) whether emergency measures were discriminatory, (2) whether countries derogated from non-derogable rights, (3) whether emergency measures failed to state a time limit, and (4) whether state security forces violated physical integrity rights while enforcing emergency measures. Similar to the Oxford data, each of these outcomes is an ordered category measuring the severity of the outcomes, including no, minor, moderate, and major violations. We do not collapse these variables into binary indicators and instead work with them as ordered categories.

@@

We use a binary indicator of whether a state derogated from the ICCPR during a given week. We coded this data from real-time state action updates recorded at the UN Treaty Collection [@UnitedNations:2023]. During the 15 months covered in this project, there were 115 formal derogations submitted mentioning 8 specific ICCPR articles; 15 derogations made no reference to any specific article. Of the derogations that indicated specific time limits, the average derogation duration was 41 days. Fifteen of the derogating states issued renewals/extensions of derogations that had time limits. We mark the start and end date of each derogation so that each country-week observation indicates whether a state has derogated.

@@

Several additional variables that are associated with and potentially confound both the decision to derogate and to enact specific policies. To account for the severity of the pandemic in each country, we control for the count of new COVID-19 cases and the count of new COVID-19 deaths, as well as the cumulative count of cases and deaths.

@@

When working with binary outcomes like the declaration of general emergency measures or the violation of non-derogable rights, we use logistic regression models; when working with ordered categorical outcomes like the degree of discriminatory policy or abusive enforcement, we use ordered logistic regression models. Because the panel data represents a complete or apparent sample of all countries [@BerkWesternWeiss:1995], we use a Bayesian approach to model the probabilities of the various outcomes and better estimate the uncertainty in the model parameters. We use weakly informative priors [@GelmanJakulinPittau:2008] for all model coefficients to keep estimates within realistic ranges and allow for reasonable uncertainty in parameter estimates.

@@

[^fixed-note]: In these formulas, we refer to all the coefficients as "population averages and coefficients." The nomenclature for these terms is inconsistent across disciplines, though, and they are also often referred to as "global effects" or "fixed effects". Broadly speaking, these terms represent the effects of each variable on the outcome across all regions or countries and do not incorporate region- or country-specific differences.

@@

We first examine how derogating and non-derogating states implemented emergency public health measures. @fig-preds-mfx-policies-1 shows the results of five logistic regression models as predicted probabilities of imposing specific emergency measures over the course of the first 15 months of the pandemic. Derogating countries are more likely to have undertook each policy, and the difference in probabilities between derogating and non-derogating states is substantial (see @fig-preds-mfx-policies-2 for the percentage-point differences between the predictions for the two groups of states; see Appendix Table 1 for numeric values). This provides strong support for both components of our first hypothesis—states that derogated were substantially more likely to impose relevant emergency policies aimed at curtailing the spread of the virus (H~1a~), while states that did not issue derogations were less likely to engage in similar policies (H~1b~).

All countries were highly likely to cancel public events in March and April 2020, with a greater than a 95% chance. Countries that did not derogate became less likely to cancel events, with the probability dropping to just above 90% in June 2021. Countries that derogated, however, regularly canceled with a nearly 100% probability. The difference in predicted probabilities between the two types of countries ranged from `r policies_mfx$cancel_public_events.min$draw_nice`–`r policies_mfx$cancel_public_events.max$draw_nice` percentage points (`r policies_mfx$cancel_public_events.max$p_gt`) between March 2020 and June 2021. The differences in policy implementation are starker for other types of emergency measures. Derogating states had a nearly 100% probability of imposing restrictions on gathering across all 15 months, while non-derogating states ranged between `r policies_pred$gathering_restrictions.No.min$draw_nice`% and `r policies_pred$gathering_restrictions.No.max$draw_nice`%, representing a `r policies_mfx$gathering_restrictions.min$draw_nice`–`r policies_mfx$gathering_restrictions.max$draw_nice` percentage point difference (`r policies_mfx$gathering_restrictions.max$p_gt`). Derogating states were also far more likely to close public transit, beginning with a `r policies_pred$close_public_transit.Yes.max$draw_nice`% probability in March 2020 and ending in June 2021 with a `r policies_pred$close_public_transit.Yes.min$draw_nice`% probability. Non-derogating states followed a similar downward trend over time, but `r policies_mfx$close_public_transit.min$draw_nice`–`r policies_mfx$close_public_transit.max$draw_nice` percentage points lower than their derogating counterparts (`r policies_mfx$close_public_transit.max$p_gt`). Derogating states were more likely to restrict movement, with a nearly 100% probability, dropping to `r policies_pred$movement.Yes.min$draw_nice`% by June 2021. Non-derogating states began the pandemic with a predicted `r policies_pred$movement.No.max$draw_nice`% probability of limiting movement, dropping to only `r policies_pred$movement.No.min$draw_nice`% by June 2021, representing a `r policies_mfx$movement.min$draw_nice`–`r policies_mfx$movement.max$draw_nice` percentage point difference (`r policies_mfx$movement.max$p_gt`). The only policy where there was no sizable difference between the two types of states is restrictions on international travel. Both derogating and non-derogating states had a nearly 100% probability of limiting international movement across all 15 months.

@@

These results show that countries implementing emergency measures took the time to uphold their treaty obligations. International law was not just an empty promise—statse that derogated were more likely to keep these emergency measures in place for longer and take their derogations seriously.
